\plaintitle

### 2020-03-12 wip-design/paper.tex
Design section

\definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
\DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
%\newcommand{\kvd}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\kvd}[1]{\textbf{#1}}
\newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont #1}}
\newcommand{\tg}{The Gamma\xspace}

@@

@@

\section{Background and design goals}
\label{sec:motivation}

@@

\includegraphics[width=1\columnwidth]{figures/lords5}

\subsection{Querying via Iterative Prompting}
The

\newpage

@@

interaction = iterative prompting
how is this different from just auto-complete?

trace = for transparency and learning

One important observation from the above list is that our tool should be accessible to non-expert
users such as readers and non-technical journalists, while providing extra capabilities for more
technical users. It should be easy to use if one just wants to modify existing code and should
encourage experimentation. Considering these challenges, we identify the following technical
design principles. todo{There must be papers on learning programming that can be referenced
here.}

\paragraph{B1. Learning from examples and by experimentation.} We should support two ways of learning.
Users of tools such as spreadsheets
often learn by looking at existing problem solutions todo{Advait's PPIG}. Our design should allow
this by making it possible to inspect and retrace steps used while solving a problem in an existing
application. Another principle of spreadsheets that we want to keep is the ability to experiment
and see results immediately. Our design should allow users to try invoking an operation or modifying
a parameter and quickly see if this leads to the desired results.

\paragraph{B2. Choice over construction.} To minimize the amount of information that users have to
learn and remember, our system should work in a way that allows constructing programs by
choosing from options that can reasonably appear in a current context, rather than requiring
users to recall particular syntax or exact identifier name.
todo{recognition over recall?}

\paragraph{B3. Make simple things easy and complex things possible.} Some users of the system may, over time, become advanced users
and the system should support those. In other words, the upper bound on what can be achieved
should be well above the most common use cases. At the same time, the complex features that power users might
need should not affect the most elementary uses of the system and should remain completely hidden
until needed. In other words, the lower bound on what one needs to know to use the system for basic
tasks should be as low as possible. todo{I think I got this idea of "boundaries" on what
is possible from some paper, but cannot recall which...}

\paragraph{B4. Visibility of state.} To support transparency, the system should make its entire state
transparent -- when reviewing a data analysis, all parameters should be immediately visible and the
user should not need to, e.g.~navigate through complex user interface to find them.

@@

\subsection{Type providers}

@@

>>>>

\definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
\DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
%\newcommand{\kvd}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\kvd}[1]{\textbf{#1}}
\newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont\small #1}}
\newcommand{\tg}{The Gamma\xspace}

@@

Both follow from assistant tools for software development \cite{assistants}. Our implementation,
outlined in the previous section, implements iterative prompting through a code editor mechanism
designed for code completion. Finally, the work on type providers \cite{inforich,fsdata,dotdriven},
which we extend, also utilizes code completion.

@@

\section{Motivation}
\label{sec:motivation}

@@

\includegraphics[width=1\columnwidth]{figures/lords5}
\vspace{-1.5em}

\subsection{Querying via Iterative Prompting}
In the example, the journalists constructs a query that filters and sorts a data table. This is
not unlike writing a query in the SQL language. However, the whole query is constructed through
a single interaction mechanism that we refer to as \emph{iterative prompting}. The user triggers
the interaction by typing `.' at the end of the query constructed so far. She is then offered a
list of available operations in the current context. This may include a choice of top-level
operations (e.g.~at the start and then again after choosing \ikvd{then}) or possible
parameters for an operation (such as possible sorting keys and directions after choosing
\ikvd{sort data}). The user then merely needs to choose one of the available options, rather than
learning the syntax and semantics of SQL in order to know what constructs can be used in the
given context.
We discuss the important aspects of The Gamma design in the next section and then discuss the
integration of specific data sources in detail, revisiting the type provider for querying
data tables.

% ==================================================================================================

@@

We identified a number of requirements for a data exploration tool for journalists earlier.
First, using the tool should result in transparent and reproducible data analyses that can be
understood and modified by non-experts. Second, the tool should minimze the number of concepts
that the user needs to understand and, subsequently, allow them to learn through exploration
and from examples. In this section, we discuss how our design aims to satisfy those requirements.

\subsection{Lowering the Barrier to Entry}
Data exploration tasks, such as querying tables, have a certain irreducible complexity.
Regardless of the interface, the user will be exposed to concepts such as filtering,
sorting and grouping. Our design aims to lower the barrier to entry by stratifying the concepts
into first-level \emph{iterative prompting} principle and second-level \emph{domain specific
languages} for each kind of data source.
The user needs to master the iterative prompting principle in order to start working
with the system. However, using individual data sources can be mastered gradually.

\paragraph{Iterative Prompting}
Iterative prompting is an interaction principle where the user repeatedly invokes an
auto-complete prompt and makes a selection from the offered options. The principle is closely
related to both code completion and the use of command line. Iterative prompting is novel as an
overarching interaction principle for program construction.

In code completion, the auto-complete
prompt is invoked only in certain contexts, e.g.~when accessing a member of an object through a
variable defined earlier in code. It requires the user to be sufficiently familiar with the
programming language in order to get to a point where they are offered a list of members. In
contrast, our system only requires choosing the initial data source. The rest of the programming
is done via iterative prompting. The iterative nature of iterative prompting makes it similar to
using the command line or REPL (read-eval-print-loop) tools. However, those repeatedly ask the
user to type code or commands.

We argue that iterative prompting is easier than other forms of interaction, because it
follows the \emph{recognition over recall} usability heuristic. The users are only required to
choose from an offered list of options, rather than having to recall a possible command
(to type in a command line) or a syntax in a text-based programming environment.

\paragraph{Domain Specific Languages}
The access to individual data sources in The Gamma is facilitated through a domain specific
language, which defines the primitives that are offered to the user in the auto-complete prompts
invoked through iterative prompting. The domain specific languages are embedded in The Gamma --
they define merely the available members (and specify how to evaluate a chain of members), but
they cannot define any custom syntax.

We discuss the individual domain specific languages for querying data tables, graph databases and
data cubes in a later section. The complexity of those languages differs. The previously discussed
language for querying tabular data is the most complex one. However, The Gamma makes it easy for
the user to start exploring and learning new languages, because they are all accessible via
iterative prompting.

\subsection{Building the Magic Escalator of Knowledge}
As discussed earlier, The Gamma is designed for users who work under tight deadlines and only
analyse data as their secondary task. The Gamma aims to support such users by having a low
barrier for entry and making it easy to learn independently.

\paragraph{Design for Percolation}
When analysing how Excel users learn Sarkar \cite{learning} points out that users learn new
features opportunistically when the usage of a feature is apparent in a spreadsheet. For
example, users can learn different functions to use in formulas, because those are visible in
the cell. Learning how to use a wizard for creating charts in this way is not possible because
it leaves no trace in the spreadsheet -- only the final result. Sarkar's recommendation is to
\emph{design for percolation}, i.e.~in a way where looking at the final result makes it apparent
what feature has been used and how.

\paragraph{Text-based Source Code}
In The Gamma, each step in the iterative prompting process results in an identifier that is
added to the source code. This means that a program constructed solely through iterative prompting
keeps a full trace of how it was created. Seeing the resulting source code provides the user all
information that they need to recreate the program, not just by copying it, but also by using the
iterative prompting mechanism. The Gamma represents code as text and allows the user to edit it
freely, so not all interactions leave a trace. For example, deleting the most recently selected
option and choosing a different one is not apparent from the result.

\subsection{Making Complex Things Possible}
Although the focus on The Gamma is simple data exploration and almost all the use cases we
envision can be achieved using the iterative prompting interaction principle, there remain a
few cases where more flexibility is needed. We aim to \emph{make simple things easy and
complex things possible}. The language thus supports a small number of additional constructs that
can only be used through text editing. The following example illustrates three additional
constructs:

\begin{thegamma}
\kvd{let} topTravel =
  expenses
    .'sort data'.'by Travel Costs descending'.then
    .paging.take(5).'get series'
    .'with key Name'.'and value Travel Costs'

charts.column(topTravel)
  .setTitle("House of Lords members by travel expenses")
  .setColors(["red","blue","green"])
\end{thegamma}

First, The Gamma allows operations with parameters such as \ikvd{take(3)} or \ikvd{setTitle("...")}.
This is currently needed when writing a query that skips or takes first N elements from a table.
The remaining features are not needed for basic data exploration, but allow other scenarios, such
as manual creation of charts. The \ikvd{let} construct can be used to define (immutable) variables
and The Gamma also supports lists written as \ikvd{[1,2,3]}.

% ==================================================================================================

@@

\subsection{Data sources}

% ==================================================================================================

@@

"i haven't done 'then' "

USABILITY ISSUES
- need to delete the rest of the line
- when you leave ".c" there, things break
- typing . without indentation does not work
- purple box vs. black wrench icons in the completion
- how do you know top level data sources
- 'games is' bit (with the year) - methods suck!
